The unambiguous and easy to interpret visualization of clinical patient records can help to facilitate decision making and exploration among physicians or clinicians scientists, especially with constantly growing clinical data sets. The MIMIC-III database is a freely accessible intensive care database, which documents in detail the course and treatment of more than 40,000 patients and was already used for visualization in the medical field. In the present work, we used Tableau Desktop, a commercial visual analytic software, to visualize MIMIC-III data by building so-called dashboards. This dashboard can be linked together and might be used as a potential visualization tool for clinical data. We developed three exemplary interactive dashboards to demonstrate three potential use cases which might be useful for health researchers and clinicians. One Dashboards gives the user a brief overview of the patient information stored in the data, another dashboard displays laboratory time series data for individual patients and with the third dashboard, a user will be able to select a patient cohort based on parameters like age, gender, diseases classification codes and others. Once created, dashboards can be adapted for specific questions by implementing dynamic parameters and thus reused. This approach might be useful to enable the interactive visualization and exploration of clinical data for non primarily technical users like health researchers and clinicians and also allows them to create customized dashboards for their own needs. These dashboards could be used as a starting point for further visualizations for the MIMIC-III Data and clinical data in general for instance in the context of a visualization platform for a data warehouse system. 