https://ieeexplore.ieee.org/document/7570894

> There are various phases in each of the analysis process which are performed in order to get the output. These phases are performed in sequential order to achieve the desired goal effectively [13]. The phases of analytics are:

Identify the problem,

Designing data requirement,

Pre-Processing data,

Performing analytics over data and e. Visualizing data.

 - 
View All

A. Identify the Problem
Nowadays, analytics are performed web datasets because of increasing use of internet and growing business of organisation over internet. This leads to gradual increase of data size day by day. Business organisation want to make predictions over the data so as to make desired decisions. Because of which analytics are performed over the datasets. Analytical applications must be scalable for collecting the datasets. Let us assume that there is an e-commerce website and wants to increase the business. So we can recognize important pages of website as high, medium and low as per popularity. We can improve business by managing web traffic and content of popular pages.

B. Designing Data Requirement
For performing the analysis process, we require datasets of the related domain. Data source can be identified on the basis of domain and problem specification. According to the problem definition, attributes of datasets are identified. For example, if we want to perform some social media analytics, so we can use data of Facebook or Twitter as datasets and profile information, likes, posts etc. can be used ad data attributes for identifying the user characteristics.

C. Pre-Processing Data
Pre-processing is done to perform operations on data so a to translate data into a specified format before providing data to algorithms or tools, data operations includes data cleaning, data aggregation, data sorting, data augmentation and data formatting. These will provide data in supported format of all tools used in analytics. Analytics process will start with this format data as input. Like, in case of big data, Hadoop Distributed File System format is used.

D. Performing Analytics Over Data
As the data is available in required format, so analytics operation can be applied to it. These operations are performed to draw out meaning information from data in order to make better decisions towards business. Analytics can be performed with various machine learning or custom algorithmic concepts like regression, classification, clustering, model-based recommendation etc. These models are further evaluated and improved by various stages of machine learning concepts. Improved or optimized algorithms can provide better results.

E. Visualizing Data
Visualization is a technique for creating images, diagrams, or animations to communicate a message. It is used for displaying output of analytics. This is used by many tools like R tool, Weka, RapidMiner etc [14].